Virtualization of a Real-Time Operating System for Robot Control with a Focus on Real-Time Compliance

### 2024-09-02 Masterarbeit_Robotik_Pamuk.tex
FINAL

\begin{itemize}
	\item This work was written at SIGMATEK GmbH \& Co KG, a company that researches, develops, produces, and distributes automation technology for industrial machinery and plant engineering. SIGMATEK uses its own custom Linux-based operating system, to be run on their self-manufactured CPUs. This operating system will be referred to as ``Salamander 4'' in this work. The details of Salamander 4 are explained in Section~\ref{sec:guest_operating_system}.
	\item Salamander 4 is virtualized through Quick Emulator (QEMU), an open-source emulator and virtualizer that allows running different operating systems on a computer. QEMU is described in Section~\ref{sec:qemu}.
	\item Salamander 4 was created with Yocto, an open-source project that helps develop custom Linux-based systems. Yocto is discussed in Section~\ref{sec:yocto}.
	\item Salamander 4 employs hard real-time with Xenomai 3, a real-time framework for Linux that enables deterministic and low-latency performance. Xenomai 3 is detailed in Section~\ref{sec:xenomai}.
	\item The goal is to virtualize Salamander 4 and approach the performance of bare metal (hardware) through real-time performance tunings, focusing on latency requirements set by SIGMATEK. These modifications are the subject of Section~\ref{sec:real-time_tunings}.
	\item The \texttt{latency} tool of the Xenomai suite to measure the scheduling latency of the real-time OS was suggested to be used by SIGMATEK. The goal was set to achieve latency values below 50 microseconds in the duration of the measurement.
	\item SIGMATEK uses the VARAN bus for communication between machines and systems, a real-time Ethernet network that was specially developed for industrial automation. Its operating principle is outlined in Section~\ref{sec:varan}.
	\item For the purpose of replacing the physical CPU, SIGMATEK provided a PCI Insert Card module PCV 522, that was plugged into the PC, as explained in Subsection~\ref{subsec:setup_experiment_virtualized}. This component could then be used to communicate with the Input/Output (I/O) peripherals, just like the physical CPU could when the I/O modules were attached to it. Subsequently, the PCV 522 module of SIGMATEK was connected with the Pulse Width Module PW 022 of SIGMATEK over the VARAN bus and the Varan Connection Module VI 021 of SIGMATEK.
	\item The robotic application from Section~\ref{sec:robotic_application} was written in Lasal Class 2, an object-oriented programming tool of SIGMATEK and part of SIGMATEK's Engineering Platform LASAL, with client-server technology and graphic representation.
\end{itemize}

@@

\section{Objective}
The goal is to bring the latency of the virtualized Salamander 4 closer to that of the hardware (bare metal), thereby providing a comprehensive blueprint for making a virtualized guest system in a host system real-time capable with deterministic behavior. Section~\ref{sec:starting_situation} outlines the initial situation of this master's thesis and serves as the foundation for the real-time performance tunings discussed in Section~\ref{sec:real-time_tunings}. The objective is to achieve latency values below 50 microseconds during the measurement period through real-time tunings, with no outliers whatsoever. To determine and validate the performance of the tunings, the \texttt{latency} tool is used, as stated in Section~\ref{sec:application_context}. On top of that, the aim is to use a robotic application in a practical scenario to demonstrate and validate that the tunings significantly reduce latency and improve determinism compared to the unmodified virtualization.

@@

\begin{figure}[H]
	\centering
	\begin{tikzpicture}[node distance = 0.35cm, auto, % Adjusted node distance
			block/.style={
					rectangle,
					draw,
					fill=orange!30,
					text width=35em,
					rounded corners,
					align=center,
					inner sep=0.5em % Adjusted inner separation
				},
			line/.style={draw, -latex'}
		]
		% Place nodes
		\node [block] (issue) {\scriptsize \textbf{Command Issuance}\\A command is sent to the system to execute a movement.};
		\node [block, below=of issue] (record1) {\scriptsize \textbf{Recording Command Time}\\The timestamp of the moment the command is issued is recorded.};
		\node [block, below=of record1] (detect) {\scriptsize \textbf{Detecting Command Arrival}\\The moment when the signal reaches the PWM is detected.};
		\node [block, below=of detect] (record2) {\scriptsize \textbf{Recording Command Arrival Time}\\The timestamp of the moment the signal reaches the PWM is recorded.};
		\node [block, below=of record2] (calculate) {\scriptsize \textbf{Calculating Command-to-PWM Latency}\\The latency is determined by the difference between the command issuance time and \\\vspace{-0.5em} the command arrival time at the PWM.};		
		% Draw edges
		\path [line] (issue) -- (record1);
		\path [line] (record1) -- (detect);
		\path [line] (detect) -- (record2);
		\path [line] (record2) -- (calculate);
	\end{tikzpicture}
	\caption[Flowchart of Robotic Application]{Flowchart of Robotic Application}
	\label{fig:robotic_application_flow}
\end{figure}

@@

\noindent The improved latency was also tested with the robotic application in Subsection~\ref{sec:robotic_application}. The difference between the command time and the time the signal reaches the PWM was measured 1,000 times to get reliable results for the bare metal, untuned, and tuned versions of Salamander 4. The results are visualized in Figure~\ref{fig:combined_latencies} and detailed in Table~\ref{tab:robotic_application_latency_values_combined}.

@@

\chapter{Discussion}\label{cha:discussion}
This master's thesis shows that the PREEMPT-RT patch along with the applied real-time performance tunings have a positive impact on reducing latency and improving determinism. The maximum latency in the final tuned version of Salamander 4 virtualization is significantly lower than in the untuned setup. The worst latency value in a 10-minute measurement span decreased from 707.622~µs to 17.134~µs after tuning BIOS, kernel, host, QEMU/KVM, and guest configurations. Especially the BIOS and kernel configurations played a crucial role in reducing the maximum latency, with a major fall to 21.694~µs. From this moment on, there were also no overruns in the test any longer. The host configurations, including CPU and interrupt affinity, and real-time prioritization of QEMU, further reduced latency down to 18.441~µs. The guest OS, Salamander 4, already had real-time capabilities through Xenomai. The focus was primarily on optimizing the host and virtualization layer to achieve the desired real-time performance. Finally, QEMU/KVM configurations, such as tuning the LAPIC timer advance and using hugepages, brought the latency down to the final worst latency value of 17.134~µs in 10 minutes. This value is very close to the bare metal value of 10.709~µs. The goal was set to achieving latency values below 50 microseconds in the duration of the measurement. This goal is achieved.

\bigskip \noindent As a next step, the robotic application in a practical scenario also demonstrates and validates the improvement of the tuned virtualization. The difference between the command time and the time the signal reaches the PWM was measured 1,000 times for the untuned, tuned, and bare metal versions of Salamander 4. The most important metric here is also the worst latency of the signal reaching the PWM. It dropped from 129~ms to 3.988~ms in the developed application. Comparing this to the worst latency value of the bare metal version, which is 1.49~ms, it is apparent that the tuned virtualization through real-time performance configurations came very close to the determinism and reliability of the hardware.

@@

\bigskip \noindent The initial latency values were first measured with the \texttt{latency} program of the Xenomai tool suite. The tests were conducted with a sampling period of 100 µs, using a periodic user-mode task, and were assigned a priority of 99. If any sample's latency value was greater than 100~µs, this was considered an overrun. There was a significant initial gap in latency statistics between the virtualized Salamander 4 and the Salamander 4 on bare metal. To address this gap, an extensive tuning process was carried out to achieve real-time performance and determinism. These modifications were added sequentially and tested together. After applying the PREEMPT-RT patch, the BIOS was configured, followed by the kernel. The BIOS settings were not reverted when moving to the kernel configurations. After these two steps, the worst latency decreased from an initial value of 707.62~µs to 21.694~µs. From this moment on, there were also no overruns in the tests any longer. The host configurations, including CPU and interrupt affinity, and real-time prioritization of QEMU, further reduced latency down to 18.441~µs. The guest OS, Salamander 4, already had real-time capabilities through Xenomai. The focus was primarily on optimizing the host and virtualization layer to achieve the desired real-time performance. Finally, QEMU/KVM configurations, such as tuning the LAPIC timer advance and using hugepages, brought the latency down a little more to the final worst latency value of 17.134~µs in 10 minutes. This value is very close to the bare metal value of 10.709~µs. The goal was set to achieving latency values below 50 microseconds in the duration of the measurement. This goal was achieved.

\bigskip \noindent On top of that, a robotic application further confirmed the improvements. The difference between the command time and the time the signal reaches the PWM was measured 1,000 times for the untuned, tuned, and bare metal versions of Salamander 4. The worst latency of the signal reaching the PWM dropped from an initial value of 129~ms to 3.988~ms. Comparing this to the worst latency value of the bare metal version, which was 1.49~ms, it is apparent that the tuned virtualization through real-time performance configurations came very close to the determinism and reliability of the hardware.

>>>>

\begin{itemize}
	\item This work was written at SIGMATEK GmbH \& Co KG, a company that researches, develops, produces, and distributes automation technology for industrial machinery and plant engineering. SIGMATEK uses its own custom Linux-based operating system, to be run on their self-manufactured CPUs. This operating system will be referred to as ``Salamander 4'' in this work. The details of Salamander 4 are explained in Section~\ref{sec:guest_operating_system}.
	\item Salamander 4 is virtualized through Quick Emulator (QEMU), an open-source emulator and virtualizer that allows running different operating systems on a computer. QEMU is described in Section~\ref{sec:qemu}.
	\item Salamander 4 was created with Yocto, an open-source project that helps develop custom Linux-based systems. Yocto is discussed in Section~\ref{sec:yocto}.
	\item Salamander 4 employs hard real-time with Xenomai 3, a real-time framework for Linux that enables deterministic and low-latency performance. Xenomai 3 is detailed in Section~\ref{sec:xenomai}.
	\item The goal is to virtualize Salamander 4 and approach the performance of bare metal (hardware) through real-time performance tunings, focusing on latency requirements set by SIGMATEK. These modifications are the subject of Section~\ref{sec:real-time_tunings}.
	\item The \texttt{latency} tool of the Xenomai suite to measure the scheduling latency of the real-time OS was suggested to be used by SIGMATEK. The goal was set to achieve latency values below 50 µs in the duration of the measurement.
	\item SIGMATEK uses the VARAN bus for communication between machines and systems, a real-time Ethernet network that was specially developed for industrial automation. Its operating principle is outlined in Section~\ref{sec:varan}.
	\item For the purpose of replacing the physical CPU, SIGMATEK provided a PCI Insert Card module PCV 522, that was plugged into the PC, as explained in Subsection~\ref{subsec:setup_experiment_virtualized}. This component could then be used to communicate with the Input/Output (I/O) peripherals, just like the physical CPU could when the I/O modules were attached to it. Subsequently, the PCV 522 module of SIGMATEK was connected with the Pulse Width Module PW 022 of SIGMATEK over the VARAN bus and the Varan Connection Module VI 021 of SIGMATEK.
	\item The robotic application from Section~\ref{sec:robotic_application} was written in Lasal Class 2, an object-oriented programming tool of SIGMATEK and part of SIGMATEK's Engineering Platform LASAL, with client-server technology and graphic representation.
\end{itemize}

@@

\section{Objective}
The goal is to bring the latency of the virtualized Salamander 4 closer to that of the hardware (bare metal), thereby providing a comprehensive blueprint for making a virtualized guest system in a host system real-time capable with deterministic behavior. Section~\ref{sec:starting_situation} outlines the initial situation of this master's thesis and serves as the foundation for the real-time performance tunings discussed in Section~\ref{sec:real-time_tunings}. The objective is to achieve latency values below 50 µs during the measurement period through real-time tunings, with no outliers whatsoever. To determine and validate the performance of the tunings, the \texttt{latency} tool is used, as stated in Section~\ref{sec:application_context}. On top of that, the aim is to use a robotic application in a practical scenario to demonstrate and validate that the tunings significantly reduce latency and improve determinism compared to the unmodified virtualization.

@@

\begin{figure}[H]
	\centering
	\begin{tikzpicture}[node distance = 0.35cm, auto, % Adjusted node distance
			block/.style={
					rectangle,
					draw,
					fill=orange!30,
					text width=35em,
					rounded corners,
					align=center,
					inner sep=0.5em % Adjusted inner separation
				},
			line/.style={draw, -latex'}
		]
		% Place nodes
		\node [block] (issue) {\scriptsize \textbf{Command Issuance}\\A command is sent to the system to execute a movement.};
		\node [block, below=of issue] (record1) {\scriptsize \textbf{Recording Command Time}\\The timestamp of the moment the command is issued is recorded.};
		\node [block, below=of record1] (detect) {\scriptsize \textbf{Detecting Command Arrival}\\The moment when the signal reaches the PWM module is detected.};
		\node [block, below=of detect] (record2) {\scriptsize \textbf{Recording Command Arrival Time}\\The timestamp of the moment the signal reaches the PWM module is recorded.};
		\node [block, below=of record2] (calculate) {\scriptsize \textbf{Calculating Command-to-PWM Latency}\\The latency is determined by the difference between the command issuance time and \\\vspace{-0.5em} the command arrival time at the PWM module.};		
		% Draw edges
		\path [line] (issue) -- (record1);
		\path [line] (record1) -- (detect);
		\path [line] (detect) -- (record2);
		\path [line] (record2) -- (calculate);
	\end{tikzpicture}
	\caption[Flowchart of Robotic Application]{Flowchart of Robotic Application}
	\label{fig:robotic_application_flow}
\end{figure}

@@

\noindent The improved latency was also tested with the robotic application in Subsection~\ref{sec:robotic_application}. The difference between the command time and the time the signal reaches the PWM module was measured 1,000 times to get reliable results for the bare metal, untuned, and tuned versions of Salamander 4. The results are visualized in Figure~\ref{fig:combined_latencies} and detailed in Table~\ref{tab:robotic_application_latency_values_combined}.

@@

\chapter{Discussion}\label{cha:discussion}
This master's thesis shows that the PREEMPT-RT patch along with the applied real-time performance tunings have a positive impact on reducing latency and improving determinism. The maximum latency in the final tuned version of Salamander 4 virtualization is significantly lower than in the untuned setup. The worst latency value in a 10-minute measurement span decreased from 707.622~µs to 17.134~µs after tuning BIOS, kernel, host, QEMU/KVM, and guest configurations. Especially the BIOS and kernel configurations played a crucial role in reducing the maximum latency, with a major fall to 21.694~µs. From this moment on, there were also no overruns in the test any longer. The host configurations, including CPU and interrupt affinity, and real-time prioritization of QEMU, further reduced latency down to 18.441~µs. The guest OS, Salamander 4, already had real-time capabilities through Xenomai. The focus was primarily on optimizing the host and virtualization layer to achieve the desired real-time performance. Finally, QEMU/KVM configurations, such as tuning the LAPIC timer advance and using hugepages, brought the latency down to the final worst latency value of 17.134~µs in 10 minutes. This value is very close to the bare metal value of 10.709~µs. The goal was set to achieving latency values below 50 µs in the duration of the measurement. This goal is achieved.

\bigskip \noindent As a next step, the robotic application in a practical scenario also demonstrates and validates the improvement of the tuned virtualization. The difference between the command time and the time the signal reaches the PWM module was measured 1,000 times for the untuned, tuned, and bare metal versions of Salamander 4. The most important metric here is also the worst latency of the signal reaching the PWM. It dropped from 129~ms to 3.988~ms in the developed application. Comparing this to the worst latency value of the bare metal version, which is 1.49~ms, it is apparent that the tuned virtualization through real-time performance configurations came very close to the determinism and reliability of the hardware.

@@

\bigskip \noindent The initial latency values were first measured with the \texttt{latency} program of the Xenomai tool suite. The tests were conducted for 10 minutes with a sampling period of 100 µs, using a periodic user-mode task, and were assigned a priority of 99. If any sample's latency value was greater than 100~µs, this was considered an overrun. There was a significant initial gap in latency statistics between the virtualized Salamander 4 and the Salamander 4 on bare metal. To address this gap, an extensive tuning process was carried out to achieve real-time performance and determinism. These modifications were added sequentially and tested together. After applying the PREEMPT-RT patch, the BIOS was configured, followed by the kernel. The BIOS settings were not reverted when moving to the kernel configurations. After these two steps, the worst latency decreased from an initial value of 707.62~µs to 21.694~µs. From this moment on, there were also no overruns in the tests any longer. The host configurations, including CPU and interrupt affinity, and real-time prioritization of QEMU, further reduced latency down to 18.441~µs. The guest OS, Salamander 4, already had real-time capabilities through Xenomai. The focus was primarily on optimizing the host and virtualization layer to achieve the desired real-time performance. Finally, QEMU/KVM configurations, such as tuning the LAPIC timer advance and using hugepages, brought the latency down a little more to the final worst latency value of 17.134~µs in 10 minutes. This value is very close to the bare metal value of 10.709~µs. The goal was set to achieving latency values below 50 µs in the duration of the measurement. This goal was achieved.

\bigskip \noindent On top of that, a robotic application further confirmed the improvements. The difference between the command time and the time the signal reaches the PWM module was measured 1,000 times for the untuned, tuned, and bare metal versions of Salamander 4. The worst latency of the signal reaching the PWM dropped from an initial value of 129~ms to 3.988~ms. Comparing this to the worst latency value of the bare metal version, which was 1.49~ms, it is apparent that the tuned virtualization through real-time performance configurations came very close to the determinism and reliability of the hardware.

### 2024-09-02 Paper.tex
FINAL

\begin{multicols}{2}
  \abstract{This thesis investigates the virtualization of a Real-Time Operating System (RTOS), which is built with Yocto, employs hard real-time through Xenomai 3, and is emulated using QEMU/KVM. The primary objective was to bridge the latency gap between virtualized and bare metal versions to ensure deterministic and reliable behavior, which is crucial for real-time robotics. Initial latency measurements showed a high gap between bare metal and virtualized setups. Extensive tuning across BIOS, kernel, host OS, QEMU/KVM, and guest OS reduced the worst-case latency from 707.622~µs to 17.134~µs, close to the bare metal performance of 10.709~µs. The improvement was validated using a robotic application, comparing tuned virtualization with untuned and hardware versions.} {Virtualization, Real-Time Systems, Latency Reduction, Robot Control}

  \section{Introduction}
  In today's industrial production and automation, robots must react to their environment and perform time-critical tasks within strict time constraints. Delays or errors can have catastrophic consequences. In a traditional general-purpose operating system (GPOS), a high-priority task cannot interrupt a kernel operation, which makes them unsuitable for real-time requirements, as they cannot guarantee deterministic execution times. However, in a real-time operating system (RTOS), a high-priority process can interrupt a lower-priority task, even if it is in the middle of a kernel operation. Virtualizing real-time operating systems for robotic control offers many advantages over traditional hardware-based approaches, especially when it comes to scaling~\cite{abbasiExploringOpenStackScalable2023}, flexibility, and costs~\cite{taccariEmbeddedRealTimeVirtualization2014}. However, virtualization also introduces overhead and latency~\cite{casiniLatencyAnalysisVirtualization2021}~\cite{zhangEvaluatingOptimizingVirtualization2010}. In this master's thesis, the goal was the virtualization of a real-time operating system to control a robot, with a focus on compliance with real-time determinism. Specifically, the aim was to bring the latency of the virtualized Salamander 4 closer to that of the bare metal version, as initial latency measurements revealed a significant latency gap between the bare metal and virtualized setup.

@@

  \section{Implementation}
  The initial latency values were first measured with the \texttt{latency} program of the Xenomai tool suite. The tests were conducted with a sampling period of 100~µs, using a periodic user-mode task, and were assigned a priority of 99. If any sample's latency value was greater than 100~µs, this was considered an overrun. There was a significant initial gap in latency statistics between the bare metal and virtualized Salamander 4. To address this gap, an extensive tuning process was carried out to achieve real-time performance and determinism. These modifications were added sequentially and tested together. Previous tunings were not reverted when moving to the next tunings. First, the BIOS was configured, followed by the kernel. The BIOS settings were not reverted when moving to the kernel configurations. Next, the host was configured, but the BIOS and kernel settings remained unchanged. The guest configurations were not changed since Xenomai already provides real-time capabilities. Finally, QEMU/KVM settings were applied. On top of that, a robotic application further confirmed the improvements. The difference between the command time and the time the signal reaches the PWM was measured 1,000 times for the untuned, tuned, and hardware versions of Salamander 4.

  \section{Results}
  This section summarizes the results after the real-time performance tunings, both in terms of the \texttt{latency} program and the robotic application. Figure~\ref{fig:max_latency_combined_results} displays the change in latency after each real-time performance tuning.

  \begin{figure}[H]
    \centering
    \includegraphics[width=1.0\columnwidth]{img/results/gnuplot_combined_max_latency_all.png}
    \caption[Comparison of Latency Distribution of Salamander 4 Configurations]{Comparison of Latency Distribution of Salamander 4 under different Configurations}
    \label{fig:max_latency_combined_results}
  \end{figure}

@@

  \begin{table}[H]
    \centering
    \footnotesize
    \caption{Comparison of Latency Statistics in Salamander 4 under different Configurations}
    \label{tab:latency_tables_combined}
    \begin{tabular}{|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{\textbf{Tuning}} & \multicolumn{3}{c|}{\textbf{Latency (µs)}} & \multirow{2}{*}{\textbf{Overruns}} \\ \cline{2-4}
      & \textbf{Min} & \textbf{Avg} & \textbf{Max} & \\ \hline
      Bare Metal & 0.613 & 1.380 & 10.709 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}Untuned\\ Virtualization\end{tabular} & 2.536 & 8.940 & 707.622 & 43 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After BIOS\\ Configurations\end{tabular} & 0.969 & 3.948 & 457.545 & 94 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After Kernel\\ Configurations\end{tabular} & 2.545 & 4.811 & 21.694 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After Host\\ Configurations\end{tabular} & 2.591 & 4.834 & 18.441 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After QEMU\\ Configurations\end{tabular} & 2.614 & 4.779 & 17.134 & 0 \\ \hline
    \end{tabular}
\end{table}

The improved latency was also tested with a robotic application. The difference between the command issuance time and the time the signal reaches the PWM was measured 1,000 times for the bare metal, untuned, and tuned versions of Salamander 4. The results are visualized in Table~\ref{tab:robotic_application_latency_values_combined}.

  \begin{table}[H]
    \centering
    \footnotesize
    \caption{Comparison of Robotic Application Latency of Salamander 4 Configurations}
    \label{tab:robotic_application_latency_values_combined}
    \begin{tabular}{|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{\textbf{Tuning}} & \multicolumn{3}{c|}{\textbf{Latency (ms)}} & \multirow{2}{*}{\begin{tabular}[c]{@{}c@{}}\textbf{Std Dev}\\ \textbf{(ms)}\end{tabular}} \\ \cline{2-4}
      & \textbf{Min} & \textbf{Avg} & \textbf{Max} & \\ \hline
      Bare Metal & 1.211  & 1.347 & 1.49 & 0.082 \\ \hline
      \begin{tabular}[c]{@{}c@{}}Untuned\\ Virtualization\end{tabular} & 3.1 & 24.603 & 129.46 & 13.876\\ \hline
      \begin{tabular}[c]{@{}c@{}}Tuned\\ Virtualization\end{tabular} & 1.219  & 2.62 & 3.988 & 0.812 \\ \hline
    \end{tabular}
\end{table}

  \section{Discussion}
  The results show that the real-time performance tunings applied to the virtualized Salamander 4 OS reduced latency and improved determinism. The maximum latency decreased from 707.622~µs to 17.134~µs after tuning BIOS, kernel, host, guest, and QEMU/KVM configurations. Especially the BIOS and kernel configurations played a crucial role in reducing the maximum latency, with a major fall to 21.694~µs. This also eliminated overruns going forward. Host configurations, including the PREEMPT-RT patch, CPU and interrupt affinity, and real-time prioritization of QEMU further reduced latency a little bit down to 18.441~µs. The guest OS already had real-time capabilities through Xenomai, hence the focus was primarily on optimizing the host and virtualization layer to achieve the desired real-time performance. Finally, QEMU/KVM configurations, such as tuning the LAPIC timer advance and using hugepages, brought the latency down to the final worst latency value of 17.134~µs. This value is very close to the bare metal value of 10.709~µs. The goal was set to achieving latency values below 50 microseconds in the duration of the measurement. This goal was achieved.
  A robotic application in a practical scenario validated the improvement of the tuned virtualization compared to the unmodified version. The difference between command time and signal reaching the PWM was measured 1,000 times for untuned, tuned, and hardware versions of Salamander 4. The worst latency dropped from 129ms to 3.988ms in the developed application. Compared to the hardware version's worst latency of 1.49ms, the tuned virtualization came very close to the determinism and reliability of the hardware.

  \section{Summary and Outlook}
  This master's thesis aimed to virtualize a real-time operating system for robot control, focusing on real-time determinism. The goal was to reduce latency in virtualized Salamander 4 to match its bare metal version. Initial tests showed a significant latency gap between virtualized and bare metal versions. Extensive tuning was performed sequentially: BIOS, kernel, host, and QEMU/KVM configurations. This process reduced worst latency from 707.62~µs to 17.134~µs, close to the bare metal value of 10.709~µs, achieving the goal of sub-50~µs latency.
  A robotic application confirmed these improvements. Measuring the time between command and PWM signal 1,000 times showed worst latency dropping from 129ms to 3.988ms, approaching the hardware version's 1.49ms.
  Future work could include further optimizations, testing other hypervisors, and evaluating performance under various workloads and stressed situations.
  Future work could include additional configurations and optimizations of the virtualization layer, investigating other hypervisors and virtualization technologies, and extensive testing under various workloads and stressed situations.

>>>>

\begin{multicols}{2}
  \abstract{This thesis investigates the virtualization of a Real-Time Operating System, which is built with Yocto, employs hard real-time through Xenomai 3, and is emulated using QEMU/KVM. The primary objective was to bridge the latency gap between virtualized and bare metal versions to ensure deterministic and reliable behavior, which is decisive for real-time robotics. Initial latency measurements showed a high gap between bare metal and virtualized setups. Extensive tunings across BIOS, kernel, host OS, QEMU/KVM, and guest OS were performed, which reduced the worst-case latency from 707.622~µs to 17.134~µs, close to the bare metal performance of 10.709~µs. The improvement was validated using a robotic application, comparing the tuned virtualization with the untuned and bare metal versions.} {Virtualization, Real-Time Systems, Latency Reduction, Robot Control}

  \section{Introduction}
  In today's industrial production and automation, robots must react to their environment and perform time-critical tasks within strict time constraints. Delays or errors can have catastrophic consequences. In a traditional General-Purpose Operating System (GPOS), a high-priority task cannot interrupt a kernel operation, which makes them unsuitable for real-time requirements, as they cannot guarantee deterministic execution times. However, in a Real-Time Operating System (RTOS), a high-priority process can interrupt a lower-priority one, even if it is in the middle of a kernel operation. Virtualizing real-time operating systems for robotic control offers many advantages over traditional hardware-based approaches, especially when it comes to scaling~\cite{abbasiExploringOpenStackScalable2023}, flexibility, and costs~\cite{taccariEmbeddedRealTimeVirtualization2014}. However, virtualization also introduces overhead and latency~\cite{casiniLatencyAnalysisVirtualization2021}~\cite{zhangEvaluatingOptimizingVirtualization2010}. In this master's thesis, the goal was the virtualization of an RTOS, namely Salamander 4, to control a robot, with a focus on compliance with real-time determinism. Specifically, the aim was to bring the latency of the virtualized Salamander 4 closer to that of the bare metal version, as initial latency measurements revealed a significant latency gap between the bare metal and virtualized setup.

@@

  \section{Practical implementation}
  The initial latency values were first measured with the \texttt{latency} program of the Xenomai tool suite. The tests were conducted for 10 minutes with a sampling period of 100~µs, using a periodic user-mode task, and were assigned a priority of 99. If any sample's latency value was greater than 100~µs, this was considered an overrun. There was a significant initial gap in latency between the bare metal and virtualized Salamander 4. To address this gap, an extensive tuning process was carried out to achieve real-time performance and determinism. These modifications were added sequentially and tested together. Previous tunings were not reverted when moving to the next tunings. First, the BIOS was configured, followed by the kernel. The BIOS settings were not reverted when moving to the kernel configurations. Next, the host was configured, but the BIOS and kernel settings remained unchanged. The guest configurations were not changed since Xenomai already provides real-time capabilities. Finally, QEMU/KVM settings were applied. On top of that, a robotic application further confirmed the improvements. The difference between the command time and the time the signal reaches the Pulse Width Modulation (PWM) Module  was measured 1,000 times to get reliable results for the bare metal, untuned, and tuned versions of Salamander 4.

  \section{Results}
  This section summarizes the results after the real-time performance tunings, both in terms of the \texttt{latency} program and the robotic application.

  \clearpage

  Figure~\ref{fig:max_latency_combined_results} displays the change in latency after each real-time performance tuning.

  \begin{figure}[H]
    \centering
    \includegraphics[width=1.0\columnwidth]{img/results/gnuplot_combined_max_latency_all.png}
    \caption{Comparison of Latency Distribution of Salamander 4 after Configurations}
    \label{fig:max_latency_combined_results}
  \end{figure}

@@

  \begin{table}[H]
    \centering
    \footnotesize
    \caption{Comparison of Latency Statistics in Salamander 4 after Configurations}
    \label{tab:latency_tables_combined}
    \begin{tabular}{|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{\textbf{Version}} & \multicolumn{3}{c|}{\textbf{Latency (µs)}} & \multirow{2}{*}{\textbf{Overruns}} \\ \cline{2-4}
      & \textbf{Min} & \textbf{Avg} & \textbf{Max} & \\ \hline
      Bare Metal & 0.613 & 1.380 & 10.709 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}Untuned\\ Virtualization\end{tabular} & 2.536 & 8.940 & 707.622 & 43 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After BIOS\\ Configurations\end{tabular} & 0.969 & 3.948 & 457.545 & 94 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After Kernel\\ Configurations\end{tabular} & 2.545 & 4.811 & 21.694 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After Host\\ Configurations\end{tabular} & 2.591 & 4.834 & 18.441 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After QEMU\\ Configurations\end{tabular} & 2.614 & 4.779 & 17.134 & 0 \\ \hline
    \end{tabular}
\end{table}

The improved latency was also tested with a robotic application. The difference between the command issuance time and the time the signal reaches the PWM module was measured 1,000 times for the bare metal, untuned, and tuned versions of Salamander 4. The results are visualized in Table~\ref{tab:robotic_application_latency_values_combined}.

  \begin{table}[H]
    \centering
    \footnotesize
    \caption{Comparison of Robotic Application Latency of Salamander 4 Configurations}
    \label{tab:robotic_application_latency_values_combined}
    \begin{tabular}{|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{\textbf{Version}} & \multicolumn{3}{c|}{\textbf{Latency (ms)}} & \multirow{2}{*}{\begin{tabular}[c]{@{}c@{}}\textbf{Std Dev}\\ \textbf{(ms)}\end{tabular}} \\ \cline{2-4}
      & \textbf{Min} & \textbf{Avg} & \textbf{Max} & \\ \hline
      Bare Metal & 1.211  & 1.347 & 1.49 & 0.082 \\ \hline
      \begin{tabular}[c]{@{}c@{}}Untuned\\ Virtualization\end{tabular} & 3.1 & 24.603 & 129.46 & 13.876\\ \hline
      \begin{tabular}[c]{@{}c@{}}Tuned\\ Virtualization\end{tabular} & 1.219  & 2.62 & 3.988 & 0.812 \\ \hline
    \end{tabular}
\end{table}

  \section{Discussion}
  The results show that the real-time performance tunings applied to the virtualized Salamander 4 OS reduced latency and improved determinism. The maximum latency decreased from 707.622~µs to 17.134~µs after tuning BIOS, kernel, host, and QEMU/KVM configurations. Especially the BIOS and kernel configurations played a crucial role in reducing the maximum latency, with a major fall to 21.694~µs. This also eliminated overruns going forward. The host configurations, including CPU and interrupt affinity, and real-time prioritization of QEMU, further reduced latency down to 18.441~µs. The guest OS already had real-time capabilities through Xenomai, hence the focus was primarily on optimizing the host and virtualization layer to achieve the desired real-time performance. Finally, QEMU/KVM configurations, such as tuning the LAPIC timer advance and using hugepages, brought the latency down to the final worst latency value of 17.134~µs. This value is very close to the bare metal value of 10.709~µs. The goal was set to achieving latency values below 50~µs in the duration of the measurement. This goal was achieved.
  A robotic application in a practical scenario validated the improvement of the tuned virtualization compared to the unmodified version. The difference between command time and signal reaching the PWM module was measured 1,000 times for the bare metal, untuned, and tuned versions of Salamander 4. The worst latency dropped from 129~ms to 3.988~ms in the developed application. Compared to the bare metal version's worst latency of 1.49~ms, the tuned virtualization came very close to the determinism and reliability of the bare metal.

  \section{Summary and Outlook}
  This master's thesis aimed to virtualize a real-time operating system for robot control, focusing on real-time determinism. The goal was to reduce latency in virtualized Salamander 4 to match its bare metal version. Initial tests showed a significant latency gap between bare metal and virtualized versions. Extensive tuning was performed sequentially, including BIOS, kernel, host, and QEMU/KVM configurations. This process reduced worst latency from 707.62~µs to 17.134~µs, close to the bare metal value of 10.709~µs, achieving the goal of sub-50~µs latency.
  A robotic application confirmed these improvements. Measuring the time between command issuance and command arrival at the PWM module 1,000 times showed worst latency dropping from 129~ms to 3.988~ms, approaching the bare metal version's 1.49~ms.  
  Future work could include additional configurations and optimizations of the virtualization layer, investigating other hypervisors and virtualization technologies, and extensive testing under various workloads and stressed situations.
